A Lyapunov Analysis of Momentum Methods \\ in Optimization

### 2016-10-24 LyapDraft.tex
fixed a few punctuation errors
cleaned the appendix a little

\begin{document}
\title{A Lyapunov Analysis of Momentum Methods \\ in  Optimization}
%\title{A Lyapunov Perspective on Momentum Methods \\ in  Optimization}
\date{}
\author{Ashia C. Wilson \and Benjamin Recht \and Michael I. Jordan}
\maketitle
\begin{abstract}
Momentum methods play a central role in convex optimization. Several momentum methods are provably optimal and all use a technique called {\em estimate sequences} to analyze their behavior.  The technique of estimate sequences has long been considered difficult to understand. In the following paper, we show there is an equivalence between the technique of estimate sequences and a family of Lyapunov functions in both continuous and discrete time. This framework allows us to develop a simplified and unified analysis of several existing algorithms, introduce a couple of interesting new algorithms. Furthermore, it strengthens the connection between algorithms and continuous time dynamical systems . 
\end{abstract}

@@

& \leq   - \dot \beta_t \left(f(X_t) - f(x^\ast) + \frac{\mu}{2}\left\|x^\ast - Z_t\right\|^2\right) \notag.

@@

%%%%%%%%%%%%%%%%%%%%FRANK WOLFE PROOF %%%%%%%%%%%%%
\subsection{Proof of Discretizations}
\label{App:OtherDiscret}
\paragraph{Method 1} We analyze~\eqref{Eq:QuasiSub2} using the Lyapunov function~\eqref{Eq:DiscLyapFunc1}. 
\begin{subequations}
\begin{align*}
E_{k+1} - E_k %&= A_{k+1}(f(x_{k+1}) -f(x^\ast)) - A_k(f(x_k) - f(x^\ast)) + D_h(x^\ast, z_{k+1}) - D_h(x^\ast, z_k) \notag\\
&= A_{k+1}(f(x_{k+1}) -f(x_k)) + \alpha_k(f(x_k) - f(x^\ast)) - \langle \nabla h(z_{k+1}) - \nabla h(z_k), x^\ast - z_{k+1} \rangle\notag\\
&\quad- D_h(z_{k+1}, z_k) \notag\\
 &\overset{\substack{\eqref{Eq:XSeqOther1}\\\eqref{Eq:ZSeqOther1}}}{=} A_{k+1}(f(x_{k+1}) -f(x_k)) + \alpha_k(f(x_k) - f(x^\ast)) + (A_{k+1} - A_k)\langle \nabla f(x_k), x^\ast - x_{k+1}\rangle  \notag\\
 &\quad -A_{k} \langle \nabla f(x_k), x_{k+1} - x_k\rangle - D_h(z_{k+1}, z_k)\\
 &= A_{k+1}(f(x_{k+1}) -f(x_k)) + \alpha_k(f(x_k) - f(x^\ast)) + (A_{k+1} - A_k)\langle \nabla f(x_k), x^\ast - x_{k}\rangle \notag\\
 &\quad  + (A_{k+1} - A_k)\langle \nabla f(x_k), x_{k}- x_{k+1}\rangle -A_{k} \langle \nabla f(x_k), x_{k+1} - x_k\rangle - D_h(z_{k+1}, z_k)\\
  &= A_{k+1}(f(x_{k+1}) -f(x_k) - \langle \nabla f(x_k), x_{k+1} - x_k\rangle) + \alpha_k(f(x_k) - f(x^\ast)+\langle \nabla f(x_k), x^\ast - x_{k}\rangle)  \notag \\ &\quad- D_h(z_{k+1}, z_k)\\
 & \leq A_{k+1} \frac{1}{2}\| x_{k+1} - x_{k}\|^2\\
% & =\frac{1}{2} \frac{A_{k+1}\alpha_k^2}{A_{k}^2} \|x_{k+1} - z_{k+1}\|^ 2\\
& \leq  \frac{1}{2} \frac{A_{k+1}\alpha_k^2}{A_{k}^2}diam(\X)^2
\end{align*}
\end{subequations}
%\end{small}
Therefore, we obtain the following upper bound, 
\begin{align*}
f(x_k) - f(x^\ast) \leq \frac{D_h(x^\ast, x_0) + A_0(f(x_0)- f(x^\ast)) +  \frac{1}{2}\sum_{i=1}^k \frac{A_{i+1}(A_{i+1} - A_{i})^2}{A_{i}^2}diam(\X)^2 }{A_k}.
\end{align*}
Choosing $A_k = \frac{k(k+1)}{2}$ gives a $O(1/k)$ convergence rate.

\paragraph{ Method 2} We analyze ~\eqref{Eq:QuasiSub3} using the Lyapunov function~\eqref{Eq:DiscLyapFunc1}. 
%\begin{small}
\begin{subequations}%\label{Eq:LyapDiscArg}
\begin{align}
%{\color{red} \frac{d}{dt}D_h(x, X_{t} + e^{-\alpha_{t}} \dot X_{t})}\,  
%& =  \int_{t_0}^{t_1} \frac{d}{ds}D_h(x, X_{s} + e^{-\alpha_{s}} \dot X_{s})\, ds +D_h(x, X_{t_0} + e^{-\alpha_{t_0}} \dot X_{t_0})\\
%& {\color{red} =  - \left\langle \frac{d}{dt} \nabla h(X_t+ e^{-\alpha_t}\dot X_t), x - X_t - e^{-\alpha_t} \dot X_t\right\rangle} \,  \notag\\
D_h(x^\ast,z_{k+1}) - D_h(x^\ast, z_k) &= -\langle \nabla h(z_{k+1}) - \nabla h(z_k), x^\ast - z_{k+1}\rangle - D_h(z_{k+1}, z_k)\notag\\
%&{\color{red} \overset{\eqref{Eq:ELBreg}}{=}e^{\alpha_t + \beta_t} \langle \nabla f(X_t), x - X_t \rangle\, - e^{ \beta_t} \langle \nabla f(X_t), \dot X\rangle}\, \notag   \\
& \overset{\eqref{Eq:XSeqMethod2}}{=} (A_{k+1} - A_k) \langle \nabla f(x_{k+1}), x^\ast - z_{k}\rangle  + (A_{k+1} - A_k) \langle \nabla f(x_{k+1}), z_{k} - z_{k+1}\rangle \notag\\
&\quad- D_h(z_{k+1}, z_k)\notag\\
& \overset{\eqref{Eq:ZSeqMethod2}}{=}(A_{k+1} - A_k) \langle \nabla f(x_{k+1}), x^\ast - x_{k+1}\rangle + A_k \langle \nabla f(x_{k+1}), x_{k+1} - x_k\rangle\notag \\&\quad+ (A_{k+1} - A_k) \langle \nabla f(x_{k+1}), z_{k} - z_{k+1}\rangle- D_h(z_{k+1}, z_k)\notag\\
& \leq(A_{k+1} - A_k) \langle \nabla f(x_{k+1}), x^\ast - x_{k+1}\rangle + A_k \langle \nabla f(x_{k+1}), x_{k+1} - x_k\rangle+\notag\\
&\quad \frac{(A_{k+1} - A_k)^2}{2} \|\nabla f(x_{k+1})\|^2 \notag\\
%
&\leq (A_{k+1} - A_k)(f(x^\ast) - f(x_{k+1})) + A_k (f(x_k) - f(x_{k+1}))\notag\\
&\quad  + \frac{(A_{k+1} - A_k)^2}{2} \|\nabla f(x_{k+1})\|^2 \notag\\
%& {\color{red}= - \frac{d}{dt}\left\{ e^{\beta_t}\left(f(X_t) - f(x)\right)\right\}\label{Eq:cond1}}\\
& = A_k (f(x_k) - f(x^\ast)) - A_{k+1}(f(x_{k+1}) - f(x^\ast))  + \frac{(A_{k+1} - A_k)^2}{2} \|\nabla f(x_{k+1})\|^2 \notag.
\end{align}
\end{subequations}
Therefore, 
\begin{align*}
E_{k+1} - E_k \leq  \frac{(A_{k+1} - A_k)^2}{2} \|\nabla f(x_{k+1})\|^2 \leq  \frac{(A_{k+1} - A_k)^2}{2} G^2
\end{align*}
and we have the following convergence rate guarantee
\begin{align*}
f(x_k) - f(x^\ast) \leq \frac{D_h(x^\ast, x_0) + A_0(f(x_0)- f(x^\ast))+ \sum_{i=1}^k\frac{(A_{i+1} - A_i)^2}{2} G^2}{A_k}
\end{align*}
which is the same convergence rate as the quasi-monotone subgradient method~\eqref{Eq:ConvQuasi}. A dual averaging term can also be added to this algorithm as well.

@@

\end{algorithm}
\noindent This algorithm satisfies the following optimality conditions,

@@

&= \left(1 - \frac{\alpha_k}{A_{k+1}} \right)\left(f(x_k) -\frac{\tilde \varepsilon_k}{A_k}\right) + \frac{\alpha_k}{A_{k+1}} f_k(x)

@@

& \overset{\eqref{Eq:Under}}{\leq}  A_k f(x_k)  - \tilde \varepsilon_k + (A_{k+1} - A_k) f(x)\

@@

A_{k+1}(f(x_{k+1}) - f(x))  &\leq A_k (f(x_k) - f(x)) + \varepsilon_{k+1}

@@

@@

&\overset{\eqref{Eq:Under}}{\leq}(1 - \tau_k) \phi_k(x)  + \tau_k f(x)

@@

f(x_{k+1}) -f(x) + \frac{\mu}{2}\|x - z_{k+1}\|^2 - \left(f(x_k) - f(x)  + \frac{\mu}{2}\|x - z_{k+1}\|^2\right) &\leq - \tau_k \left(f(x_k) - f(x)  + \frac{\mu}{2}\|x - z_{k+1}\|^2\right) 

@@

Going the other direction, we have 

@@

A_{k+1} \phi_{k+1} &\leq A_k \phi_k + (A_{k+1} - A_k) f(x)
Summing over the righthand side, we obtain the estimate sequence~\eqref{Eq:Ineq1} 

@@

& =  \Big(1 -\frac{1}{A_{k+1}}\Big) f(x) + \frac{1}{A_{k+1}} \phi_0(x)\\

>>>>

\begin{document}
\title{A Lyapunov Analysis of Momentum Methods \\ in  Optimization}
%\title{A Lyapunov Perspective on Momentum Methods \\ in  Optimization}
\date{}
\author{Ashia C. Wilson \and Benjamin Recht \and Michael I. Jordan}
\maketitle
\begin{abstract}
Momentum methods play a central role in convex optimization. Several momentum methods are provably optimal and all use a technique called {\em estimate sequences} to analyze their behavior.  The technique of estimate sequences has long been considered difficult to understand. In the following paper, we show there is an equivalence between the technique of estimate sequences and a family of Lyapunov functions in both continuous and discrete time. This framework allows us to develop a simplified and unified analysis of several existing algorithms and introduce a couple of new algorithms. Furthermore, it provides insights into the art of algorithm design by strengthening the connection between algorithms and continuous time dynamical systems. 
\end{abstract}

@@

& \leq   - \dot \beta_t \left(f(X_t) - f(x^\ast) + \frac{\mu}{2}\left\|x^\ast - Z_t\right\|^2\right) \notag,

@@

%%%%%%%%%%%%%%%%%%%%FRANK WOLFE PROOF %%%%%%%%%%%%%
\subsection{Proof of Discretizations}
\label{App:OtherDiscret}
\paragraph{Method 1} We analyze~\eqref{Eq:QuasiSub2} using the Lyapunov function~\eqref{Eq:DiscLyapFunc1}. 
\begin{subequations}
\begin{align}
E_{k+1} - E_k %&= A_{k+1}(f(x_{k+1}) -f(x^\ast)) - A_k(f(x_k) - f(x^\ast)) + D_h(x^\ast, z_{k+1}) - D_h(x^\ast, z_k) \notag\\
&= A_{k+1}(f(x_{k+1}) -f(x_k)) + \alpha_k(f(x_k) - f(x^\ast)) - \langle \nabla h(z_{k+1}) - \nabla h(z_k), x^\ast - z_{k+1} \rangle\notag\\
&\quad- D_h(z_{k+1}, z_k) \notag\\
 &= A_{k+1}(f(x_{k+1}) -f(x_k)) + \alpha_k(f(x_k) - f(x^\ast)) + (A_{k+1} - A_k)\langle \nabla f(x_k), x^\ast - x_{k+1}\rangle \notag\\
 &\quad -A_{k} \langle \nabla f(x_k), x_{k+1} - x_k\rangle - D_h(z_{k+1}, z_k) \label{Eq:Stuff}\\
 &= A_{k+1}(f(x_{k+1}) -f(x_k)) + \alpha_k(f(x_k) - f(x^\ast)) + (A_{k+1} - A_k)\langle \nabla f(x_k), x^\ast - x_{k}\rangle \notag\\
 &\quad  + (A_{k+1} - A_k)\langle \nabla f(x_k), x_{k}- x_{k+1}\rangle -A_{k} \langle \nabla f(x_k), x_{k+1} - x_k\rangle - D_h(z_{k+1}, z_k)\notag\\
  &= A_{k+1}(f(x_{k+1}) -f(x_k) - \langle \nabla f(x_k), x_{k+1} - x_k\rangle) + \alpha_k(f(x_k) - f(x^\ast)+\langle \nabla f(x_k), x^\ast - x_{k}\rangle)  \notag \\ &\quad- D_h(z_{k+1}, z_k) \notag\\
 & \leq A_{k+1} \frac{1}{2}\| x_{k+1} - x_{k}\|^2 \notag\\
% & =\frac{1}{2} \frac{A_{k+1}\alpha_k^2}{A_{k}^2} \|x_{k+1} - z_{k+1}\|^ 2\\
& \overset{\eqref{Eq:XSeqOther1}}{\leq}  \frac{1}{2} \frac{A_{k+1}\alpha_k^2}{A_{k}^2}diam(\X)^2
\end{align}
\end{subequations}
%\end{small}
where in~\eqref{Eq:Stuff} we used~\eqref{Eq:XSeqOther1} and~\eqref{Eq:ZSeqOther1}. Therefore, we obtain the following upper bound, 
\begin{align*}
f(x_k) - f(x^\ast) \leq \frac{D_h(x^\ast, x_0) + A_0(f(x_0)- f(x^\ast)) +  \frac{1}{2}\sum_{i=1}^k \frac{A_{i+1}(A_{i+1} - A_{i})^2}{A_{i}^2}diam(\X)^2 }{A_k}.
\end{align*}
Choosing $A_k = \frac{k(k+1)}{2}$ gives a $O(1/k)$ convergence rate.

\paragraph{ Method 2} We analyze ~\eqref{Eq:QuasiSub3} using the Lyapunov function~\eqref{Eq:DiscLyapFunc1}. Denoting $D^\ast_{k+1, k} = D_h(x^\ast,z_{k+1})- D_h(x^\ast,z_{k})$, observe that
%\begin{small}
\begin{subequations}%\label{Eq:LyapDiscArg}
\begin{align}
%{\color{red} \frac{d}{dt}D_h(x, X_{t} + e^{-\alpha_{t}} \dot X_{t})}\,  
%& =  \int_{t_0}^{t_1} \frac{d}{ds}D_h(x, X_{s} + e^{-\alpha_{s}} \dot X_{s})\, ds +D_h(x, X_{t_0} + e^{-\alpha_{t_0}} \dot X_{t_0})\\
%& {\color{red} =  - \left\langle \frac{d}{dt} \nabla h(X_t+ e^{-\alpha_t}\dot X_t), x - X_t - e^{-\alpha_t} \dot X_t\right\rangle} \,  \notag\\
D^\ast_{k+1, k} &= -\langle \nabla h(z_{k+1}) - \nabla h(z_k), x^\ast - z_{k+1}\rangle - D_h(z_{k+1}, z_k)\notag\\
%&{\color{red} \overset{\eqref{Eq:ELBreg}}{=}e^{\alpha_t + \beta_t} \langle \nabla f(X_t), x - X_t \rangle\, - e^{ \beta_t} \langle \nabla f(X_t), \dot X\rangle}\, \notag   \\
& \overset{\eqref{Eq:XSeqMethod2}}{=} (A_{k+1} - A_k) \langle \nabla f(x_{k+1}), x^\ast - z_{k}\rangle  + (A_{k+1} - A_k) \langle \nabla f(x_{k+1}), z_{k} - z_{k+1}\rangle - D_h(z_{k+1}, z_k)\notag\\
& \overset{\eqref{Eq:ZSeqMethod2}}{=}(A_{k+1} - A_k) \langle \nabla f(x_{k+1}), x^\ast - x_{k+1}\rangle + A_k \langle \nabla f(x_{k+1}), x_{k+1} - x_k\rangle\notag \\&\quad- D_h(z_{k+1}, z_k)\notag\\
& \leq(A_{k+1} - A_k) \langle \nabla f(x_{k+1}), x^\ast - x_{k+1}\rangle + A_k \langle \nabla f(x_{k+1}), x_{k+1} - x_k\rangle+ \frac{(A_{k+1} - A_k)^2}{2} \|\nabla f(x_{k+1})\|^2 \notag\\
%
&\leq (A_{k+1} - A_k)(f(x^\ast) - f(x_{k+1})) + A_k (f(x_k) - f(x_{k+1}))  + \frac{(A_{k+1} - A_k)^2}{2} \|\nabla f(x_{k+1})\|^2 \notag\\
%& {\color{red}= - \frac{d}{dt}\left\{ e^{\beta_t}\left(f(X_t) - f(x)\right)\right\}\label{Eq:cond1}}\\
& = A_k (f(x_k) - f(x^\ast)) - A_{k+1}(f(x_{k+1}) - f(x^\ast))  + \frac{(A_{k+1} - A_k)^2}{2} \|\nabla f(x_{k+1})\|^2 \notag.
\end{align}
\end{subequations}
Therefore, 
\begin{align*}
E_{k+1} - E_k \leq  \frac{(A_{k+1} - A_k)^2}{2} \|\nabla f(x_{k+1})\|^2 \leq  \frac{(A_{k+1} - A_k)^2}{2} G^2
\end{align*}
and we have the following convergence rate guarantee
\begin{align*}
f(x_k) - f(x^\ast) \leq \frac{D_h(x^\ast, x_0) + A_0(f(x_0)- f(x^\ast))+ \sum_{i=1}^k\frac{(A_{i+1} - A_i)^2}{2} G^2}{A_k}
\end{align*}
which is the same convergence rate as the quasi-monotone subgradient method~\eqref{Eq:ConvQuasi}. A dual averaging term can also be added to this algorithm as well.

@@

\noindent This algorithm satisfies the following optimality conditions, 

@@

&= \left(1 - \frac{\alpha_k}{A_{k+1}} \right)\left(f(x_k) -\frac{\tilde \varepsilon_k}{A_k}\right) + \frac{\alpha_k}{A_{k+1}} f_k(x).

@@

& \overset{\eqref{Eq:Under}}{\leq}  A_k f(x_k)  - \tilde \varepsilon_k + (A_{k+1} - A_k) f(x).\

@@

A_{k+1}(f(x_{k+1}) - f(x))  &\leq A_k (f(x_k) - f(x)) + \varepsilon_{k+1},

@@

A_k \phi_k(x) &\leq (A_k -A_0) f(x) + A_0 \phi_0(x) + \tilde \varepsilon_{k+1}\label{Eq:EstSeq}.

@@

&\overset{\eqref{Eq:Under}}{\leq}(1 - \tau_k) \phi_k(x)  + \tau_k f(x).

@@

f(x_{k+1}) -f(x) + \frac{\mu}{2}\|x - z_{k+1}\|^2 - \left(f(x_k) - f(x)  + \frac{\mu}{2}\|x - z_{k+1}\|^2\right) &\leq - \tau_k \left(f(x_k) - f(x)  + \frac{\mu}{2}\|x - z_{k+1}\|^2\right). 

@@

Going the other direction, we have, 

@@

A_{k+1} \phi_{k+1} &\leq A_k \phi_k + (A_{k+1} - A_k) f(x).
Summing over the righthand side, we obtain the estimate sequence~\eqref{Eq:Ineq1}: 

@@

& =  \Big(1 -\frac{1}{A_{k+1}}\Big) f(x) + \frac{1}{A_{k+1}} \phi_0(x).\\
